\documentclass[aspectratio=169, handout, t]{beamer}

%\usepackage[table]{xcolor}
\mode<presentation> {
\setbeamercovered{transparent}
  \usetheme{Boadilla}

\renewcommand{\familydefault}{cmss}
\usepackage{bm}
\usepackage{listings}
\useinnertheme{rectangles}
}
\usepackage{amsmath}
\usepackage{bbold}
\usepackage{tcolorbox}
% Fix missing bold-italic sans-serif font shapes
\DeclareFontShape{OT1}{cmss}{bx}{it}{<->ssub*cmss/bx/sl}{}
\DeclareFontShape{OT1}{cmss}{b}{it}{<->ssub*cmss/bx/sl}{}
\setbeamercolor{normal text}{fg=black}
\setbeamercolor{structure}{fg= blue}
\definecolor{trial}{cmyk}{1,0,0, 0}
\definecolor{trial2}{cmyk}{0.00,0,1, 0}
\definecolor{darkgreen}{rgb}{0,.4, 0.1}
\definecolor{darkpurple}{rgb}{0.4, 0, 0.6}
\usepackage{array}
\beamertemplatesolidbackgroundcolor{white}  \setbeamercolor{alerted
text}{fg=darkpurple}
\setbeamertemplate{caption}[numbered]\newcounter{mylastframe}

\font\domino=domino
\def\die#1{{\domino#1}}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{colortbl}

\renewcommand{\familydefault}{cmss}

\usepackage{tikz}
\usepackage{lipsum}
\usepackage{booktabs}

\lstset{%
  language=R,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{darkgreen},
  stringstyle=\color{darkpurple},
  showstringspaces=false,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!10}
}

 \newenvironment{changemargin}[3]{%
 \begin{list}{}{%
 \setlength{\topsep}{0pt}%
 \setlength{\leftmargin}{#1}%
 \setlength{\rightmargin}{#2}%
 \setlength{\topmargin}{#3}%
 \setlength{\listparindent}{\parindent}%
 \setlength{\itemindent}{\parindent}%
 \setlength{\parsep}{\parskip}%
 }%
\item[]}{\end{list}}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta, positioning}
\usepackage{pgfplots}
\pgfplotsset{compat=1.17}
\usecolortheme{lily}

\newtheorem{com}{Comment}
\newtheorem{lem} {Lemma}
\newtheorem{prop}{Proposition}
\newtheorem{condition}{Condition}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem{cor}{Corollary}
\newtheorem{obs}{Observation}
 \numberwithin{equation}{section}

\makeatletter
\def\beamerorig@set@color{%
  \pdfliteral{\current@color}%
  \aftergroup\reset@color
}
\def\beamerorig@reset@color{\pdfliteral{\current@color}}
\makeatother
\setbeamertemplate{navigation symbols}{}

\useoutertheme{miniframes}
\title[PLSC 30700]{Linear Models Lecture 15: GMM Estimation and Efficiency}

\author{Robert Gulotty}
\institute[Chicago]{University of Chicago}
\vspace{0.3in}


\begin{document}

%========================================================
% SECTION 1: MOTIVATION
%========================================================
\section{Motivation}

%--- Slide 1: Title ---
\begin{frame}
\titlepage
\end{frame}

%--- Slide 2: Where We Left Off ---
\begin{frame}{Where We Left Off}

In Lecture 14, we introduced IV estimation as a \alert{method of moments}:
$$\frac{1}{n}\sum_{i=1}^n Z_i(Y_i - X_i'\beta) = 0$$

\medskip
This works when the number of moment conditions $\ell$ equals the number of parameters $k$.

\medskip
\alert{But what if we have more instruments than parameters?}

\begin{itemize}
\item We cannot simultaneously set all moment conditions to zero
\item We need a principled way to \textit{get as close as possible}
\item This is the \textbf{overidentification problem}
\end{itemize}

\end{frame}

%--- Slide 3: The Overidentification Problem ---
\begin{frame}{The Problem}

Suppose you have $\ell > k$ moment conditions:
$$\mathbb{E}[g_i(\beta)] = 0, \qquad g_i \in \mathbb{R}^\ell, \quad \beta \in \mathbb{R}^k$$

\pause
In a sample, you cannot set all $\ell$ conditions to zero simultaneously---there are more equations than unknowns.

\pause
\medskip
\textbf{Example:} You have three valid instruments for one endogenous variable.  Each implies $\mathbb{E}[Z_{ji} e_i] = 0$.  No single $\hat\beta$ satisfies all three exactly.

\pause
\medskip
We need a principled way to get \emph{as close as possible} to satisfying all of them.  The choice of ``close'' will determine both the estimator and its efficiency.

\end{frame}

%--- Slide 4: Moment Equation Models ---
\begin{frame}{Moment Equation Models}

Let $g_i(\beta)$ be a known $\ell \times 1$ function of the $i^{th}$ observation and a $k \times 1$ parameter $\beta$.

\begin{defn}
A \textbf{moment equation model} is defined by
$$\mathbb{E}[g_i(\beta)] = 0$$
and a parameter space $\beta \in B$.
\end{defn}

\medskip
\textbf{Identification requires $\ell \geq k$:}
\begin{itemize}
\item $\ell = k$: \textbf{Just identified} --- exactly enough information
\item $\ell > k$: \textbf{Overidentified} --- excess information (testable restrictions)
\item $\ell < k$: \textbf{Underidentified} --- insufficient information
\end{itemize}

\smallskip
\textbf{Example:} IV model: $g_i(\beta) = Z_i(Y_i - X_i'\beta)$

\end{frame}

%--- Slide 5: Everything Is a Moment Condition ---
\begin{frame}{Everything Is a Moment Condition}

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
\textbf{Estimator} & \textbf{Moment condition $g_i(\beta)$} & \textbf{MME} \\
\midrule
Sample mean & $Y_i - \mu$ & $\hat\mu = \bar{Y}$ \\[4pt]
OLS & $X_i(Y_i - X_i'\beta)$ & $\hat\beta = (X'X)^{-1}X'Y$ \\[4pt]
IV & $Z_i(Y_i - X_i'\beta)$ & $\hat\beta = (Z'X)^{-1}Z'Y$ \\[4pt]
2SLS & $Z_i(Y_i - X_i'\beta)$ with $W=(Z'Z)^{-1}$ & $\hat\beta_{2SLS}$ \\[4pt]
GLS & $\bar X_i\Sigma^{-1}(Y_i - \bar X_i'\beta)$ & $\hat\beta_{GLS}$ \\
\bottomrule
\end{tabular}
\end{table}

\medskip
\begin{tcolorbox}[colback=blue!5, colframe=blue!50]
\textbf{Key insight:} All the estimators we have studied are \textit{method of moments estimators} --- they solve $\bar{g}_n(\hat\beta) = 0$. GMM generalizes to the overidentified case.
\end{tcolorbox}

\end{frame}

%--- Slide: Where Do Moment Conditions Come From? ---
\begin{frame}{Where Do Moment Conditions Come From?}

The table shows that every estimator \emph{solves} moment conditions.  But where do the conditions come from?

\medskip\pause
They come from \textbf{substantive claims about the world}---not from algebra.

\medskip\pause
A theory says something about what is exogenous, what is excludable, or where randomization occurs.  That claim translates into an orthogonality condition $\mathbb{E}[g(W_i,\theta)]=0$.  The estimator follows.

\medskip\pause
\alert{The hard part is the translation.}  The same research question can lead to different moment conditions depending on what you are willing to assume.

\medskip
We'll see this with two examples.

\end{frame}

%--- Slide: Example 1 Setup ---
\begin{frame}{Example 1: Does Media Affect Political Attitudes?}

Kern \& Hainmueller (2009) study whether exposure to West German television reduced support for the East German regime.

\medskip\pause
\textbf{The question:} Does media exposure ($X$) affect regime support ($Y$)?

\medskip\pause
\textbf{The problem:} People who watch more television may differ from those who don't.  If media consumption reflects pre-existing dissatisfaction, OLS confounds the effect of media with the effect of prior attitudes.

\medskip\pause
What moment condition should we write down?  That depends on what we think is generating the variation in $X$.

\end{frame}

%--- Slide: Example 1 Two Claims ---
\begin{frame}{Two Claims, Two Moment Conditions}

\textbf{Claim A:} Media consumption is determined by geography (terrain blocks the signal).  After controlling for observables, exposure is unrelated to unobserved attitudes.
$$\mathbb{E}[X_i \, e_i] = 0 \qquad \Longrightarrow \qquad \text{OLS}$$

\pause
\textbf{Claim B:} Even geographic variation is suspect---people in ``no signal'' areas may differ in other ways.  But signal availability ($Z$) shifts exposure without directly affecting attitudes.
$$\mathbb{E}[Z_i \, e_i] = 0 \qquad \Longrightarrow \qquad \text{IV}$$

\pause
Kern \& Hainmueller make Claim B.  The instrument is geographic signal availability.  \alert{The theory determines the moment condition.  The moment condition determines the estimator.}

\end{frame}

%--- Slide: Example 2 Setup ---
\begin{frame}{Example 2: Does Winning Office Help You Win Again?}

Lee (2008) studies the \textbf{incumbency advantage}: does holding office give you an electoral edge next time?

\medskip\pause
\textbf{The problem:} Winners and losers differ in quality, funding, and district partisanship.  The moment condition $\mathbb{E}[D_i \, e_i] = 0$ (where $D_i = $ won) is not credible.

\medskip\pause
\textbf{The design:} In very close elections, who wins is approximately random.  A candidate who wins with 50.1\% is comparable to one who loses with 49.9\%.

\medskip\pause
This is a \textbf{local} claim: the theory doesn't say winning is random everywhere---only near the threshold.  The moment condition must reflect this.

\end{frame}

%--- Slide: Example 2 Local Moment Condition ---
\begin{frame}{A Local Claim $\Rightarrow$ A Local Moment Condition}

In a regression discontinuity design, we estimate the treatment effect at the cutoff $c$ using observations nearby.

\medskip\pause
Local linear regression solves a \textbf{kernel-weighted} moment condition:
$$\mathbb{E}\!\left[K_h(X_i - c) \cdot
\begin{pmatrix} 1 \\ X_i - c \end{pmatrix}
\!\left(Y_i - \alpha - \tau D_i - \beta(X_i - c)\right)\right] = 0$$

where $K_h$ is a kernel that downweights observations far from $c$.

\medskip\pause
This is recognizably a weighted OLS moment condition.  But the \textbf{kernel weight encodes the theoretical claim}: only observations near the cutoff satisfy the ``as good as random'' condition.

\medskip\pause
The bandwidth $h$ reflects how far from the cutoff you trust the continuity assumption.  Wider $h$: more data, more precision, but the claim is more fragile.  Narrower $h$: more credible, but noisier.

\end{frame}

%--- Slide: Synthesis ---
\begin{frame}{Three Claims, Three Moment Conditions}

\renewcommand{\arraystretch}{1.6}
\begin{tabular}{lll}
\toprule
\textbf{Theoretical claim} & \textbf{Moment condition} & \textbf{Estimator} \\
\midrule
Treatment is randomly assigned & $\mathbb{E}[D_i \, e_i] = 0$ & OLS \\
Instrument is excludable & $\mathbb{E}[Z_i \, e_i] = 0$ & IV / 2SLS \\
Local randomization at cutoff & $\mathbb{E}[K_h(X_i) \, D_i \, e_i] = 0$ & Local linear RD \\
\bottomrule
\end{tabular}
\renewcommand{\arraystretch}{1.0}

\medskip\pause
Each row reflects a \emph{different} claim about where the exogenous variation comes from.  Stronger claims (top) use more data.  Weaker claims (bottom) use less, but are more credible.

\medskip\pause
The moment condition framework makes these tradeoffs explicit.  When you have \emph{more} moment conditions than parameters---as with multiple instruments---GMM tells you how to combine them optimally and how to test whether they are mutually consistent.

\end{frame}

%========================================================
% SECTION 2: THE GMM ESTIMATOR
%========================================================
\section{The GMM Estimator}

%--- Slide 6: The Overidentification Problem ---
\begin{frame}{Getting as Close as Possible}

Define the sample moment:
$$\bar{g}_n(\beta) = \frac{1}{n}\sum_{i=1}^n g_i(\beta)$$

\smallskip
When $\ell = k$: We can solve $\bar{g}_n(\hat\beta) = 0$ exactly.

\smallskip
When $\ell > k$: There are \alert{more equations than unknowns}. In general, no $\beta$ sets $\bar{g}_n(\beta) = 0$.

\medskip
\textbf{Intuition:} Think of $\mu = Z'Y$, $G = Z'X$, and $\eta = \mu - G\beta$. We want $\eta$ small. Regressing $\mu$ on $G$:
$$\tilde\beta = (G'G)^{-1}G'\mu$$
minimizes $\eta'\eta$. But we can do \textit{better} with weighted least squares...

\end{frame}

%--- Slide 7: The GMM Criterion Function ---
\begin{frame}{The GMM Criterion Function}

For a positive definite $\ell \times \ell$ weight matrix $W$, the \textbf{GMM criterion function} is:

$$\boxed{J(\beta) = n\,\bar{g}_n(\beta)' W\, \bar{g}_n(\beta)}$$

\medskip
\begin{itemize}
\item $J(\beta) \geq 0$ for all $\beta$ (since $W > 0$)
\item When $W = I_\ell$: $J(\beta) = n\|\bar{g}_n(\beta)\|^2$ (Euclidean distance)
\item The factor $n$ is for distributional convenience
\item \alert{Different choices of $W$ yield different estimators} --- and different efficiency
\end{itemize}

\smallskip
For the linear IV model:
$$J(\beta) = n(Z'Y - Z'X\beta)' W (Z'Y - Z'X\beta)$$

\end{frame}

%--- Slide 8: The GMM Estimator ---
\begin{frame}{The GMM Estimator}

\begin{defn}[GMM Estimator]
$$\hat\beta_{\text{gmm}} = \arg\min_\beta J(\beta) = \arg\min_\beta\; n\,\bar{g}_n(\beta)' W\, \bar{g}_n(\beta)$$
\end{defn}

\bigskip
\begin{thm}[13.1]
For the overidentified linear IV model, the GMM estimator is:
$$\hat\beta_{\text{gmm}} = (X'ZWZ'X)^{-1}(X'ZWZ'Y)$$
\end{thm}

\bigskip
\begin{itemize}
\item When $\ell = k$ (just identified): $\hat\beta_{\text{gmm}} = (Z'X)^{-1}(Z'Y) = \hat\beta_{\text{iv}}$, regardless of $W$
\item When $\ell > k$: the estimator \alert{depends on $W$}
\end{itemize}

\end{frame}

%--- Slide 9: 2SLS as GMM ---
\begin{frame}{2SLS as GMM}

\begin{thm}[13.2]
If $W = (Z'Z)^{-1}$, then $\hat\beta_{\text{gmm}} = \hat\beta_{\text{2sls}}$.

Furthermore, if $k = \ell$ then $\hat\beta_{\text{gmm}} = \hat\beta_{\text{iv}}$.
\end{thm}

\medskip
\textbf{Proof sketch:} Substitute $W = (Z'Z)^{-1}$ into Theorem 13.1:
\begin{align*}
\hat\beta &= (X'Z(Z'Z)^{-1}Z'X)^{-1}(X'Z(Z'Z)^{-1}Z'Y) = (X'P_ZX)^{-1}(X'P_ZY)
\end{align*}
where $P_Z = Z(Z'Z)^{-1}Z'$ is the projection matrix --- exactly the 2SLS formula.

\medskip
\begin{tcolorbox}[colback=blue!5, colframe=blue!50]
\textbf{Takeaway:} 2SLS is a \textit{one-step} GMM estimator. But $(Z'Z)^{-1}$ is not always the best weight matrix.
\end{tcolorbox}

\end{frame}

%--- Slide 10: Asymptotic Distribution ---
\begin{frame}{Asymptotic Distribution of GMM}

Let $Q = \mathbb{E}[ZX']$ and $\Omega = \mathbb{E}[ZZ'e^2]$.

\begin{thm}[13.3: Asymptotic Distribution of GMM]
As $n \to \infty$,
$$\sqrt{n}(\hat\beta_{\text{gmm}} - \beta) \xrightarrow{d} N(0, V_\beta)$$
where
$$\boxed{V_\beta = (Q'WQ)^{-1}(Q'W\Omega WQ)(Q'WQ)^{-1}}$$
\end{thm}

\bigskip
\begin{itemize}
\item This is a \textbf{sandwich form}: ``bread'' $(Q'WQ)^{-1}$ wraps ``meat'' $Q'W\Omega WQ$
\item Applies for \textit{any} positive definite weight matrix $W$
\item Different $W$ $\Rightarrow$ different $V_\beta$ $\Rightarrow$ different efficiency
\end{itemize}

\end{frame}

%--- Slide 11: What Is the Best Weighting? ---
\begin{frame}{What Is the Best Weighting?}

\textbf{Question:} Which $W$ minimizes $V_\beta$?

\bigskip
\textbf{Intuition:} We want to weight moments that are
\begin{itemize}
\item \textit{more informative} (high signal) $\Rightarrow$ weight up
\item \textit{less noisy} (low variance) $\Rightarrow$ weight up
\end{itemize}

\bigskip
The variance of the sample moments is $\Omega = \mathbb{E}[g_ig_i']$. So \alert{weight inversely to their variance}: $W = \Omega^{-1}$.

\bigskip
\textbf{Analogy:} Just like GLS weights observations by the inverse of $\text{Var}(e_i)$, efficient GMM weights \textit{moment conditions} by the inverse of their covariance.

\end{frame}

%--- Slide 12: Efficient GMM ---
\begin{frame}{Efficient GMM}

\begin{thm}[13.4: Efficient GMM]
Setting $W = \Omega^{-1}$, as $n\to\infty$:
$\sqrt{n}(\hat\beta_{\text{gmm}} - \beta) \xrightarrow{d} N(0, V_\beta)$ where $V_\beta = (Q'\Omega^{-1}Q)^{-1}$.
\end{thm}

The sandwich \alert{collapses}: $(Q'\Omega^{-1}Q)^{-1}(\underbrace{Q'\Omega^{-1}\Omega \Omega^{-1}Q}_{Q'\Omega^{-1}Q})(Q'\Omega^{-1}Q)^{-1}$

\begin{thm}[13.5: Efficiency]
For any $W > 0$:
$(Q'WQ)^{-1}(Q'W\Omega WQ)(Q'WQ)^{-1} - (Q'\Omega^{-1}Q)^{-1} \geq 0$.
\end{thm}

No GMM estimator with these moment conditions can have a smaller asymptotic variance.  Chamberlain (1987): this is the \textbf{semiparametric efficiency bound}.

\end{frame}

%========================================================
% SECTION 3: 2SLS vs. EFFICIENT GMM
%========================================================
\section{2SLS vs.\ Efficient GMM}

%--- Slide 13: When Is 2SLS Efficient? ---
\begin{frame}{When Is 2SLS Efficient?}

Recall 2SLS uses $W = (Z'Z)^{-1}$, which converges to $(\mathbb{E}[ZZ'])^{-1}$.

\medskip
The efficient weight is $\Omega^{-1} = (\mathbb{E}[ZZ'e^2])^{-1}$.

\medskip
These are equal when $\mathbb{E}[e^2 | Z] = \sigma^2$ (conditional homoskedasticity):
$$\mathbb{E}[ZZ'e^2] = \sigma^2 \mathbb{E}[ZZ']$$

\begin{thm}[13.6]
Under conditional homoskedasticity $\mathbb{E}[e^2 | Z] = \sigma^2$, the 2SLS estimator $\hat\beta_{\text{2sls}}$ is efficient GMM.
\end{thm}

\smallskip
\alert{Implications:}
\begin{itemize}
\item Homoskedastic errors $\Rightarrow$ 2SLS is fine, no need for GMM
\item Heteroskedastic errors $\Rightarrow$ 2SLS is \textit{inefficient}; use efficient GMM
\end{itemize}

\end{frame}

%--- Slide 14: Two-Step GMM ---
\begin{frame}{Two-Step GMM}

\textbf{Problem:} $\Omega = \mathbb{E}[ZZ'e^2]$ is unknown. We need to estimate it.

\medskip
\textbf{Two-Step GMM:}
\begin{enumerate}
\item \textbf{Step 1:} Estimate $\beta$ by 2SLS (using $W = (Z'Z)^{-1}$). Compute residuals $\tilde{e}_i = Y_i - X_i'\hat\beta_{\text{2sls}}$.
\item \textbf{Step 2:} Estimate $\hat\Omega = \frac{1}{n}\sum_{i=1}^n Z_iZ_i'\tilde{e}_i^2$, set $\hat{W} = \hat\Omega^{-1}$, and re-estimate $\beta$.
\end{enumerate}

\begin{thm}[13.7]
The two-step GMM estimator is asymptotically efficient:
$$\sqrt{n}(\hat\beta_{\text{gmm}} - \beta) \xrightarrow{d} N(0, V_\beta)\quad\text{where}\quad V_\beta = (Q'\Omega^{-1}Q)^{-1}$$
\end{thm}

\textbf{Key point:} The initial estimator does not affect the asymptotic distribution.

\end{frame}

%--- Slide 15: Iterated GMM ---
\begin{frame}{Iterated GMM}

The two-step estimator's \textit{finite-sample} performance can depend on the initial estimator. To remove this dependence:

\medskip
\textbf{Iterated GMM:}
\begin{enumerate}
\item Start with $\hat\beta^{(0)}$ (e.g., 2SLS)
\item Compute $\hat\Omega^{(s)} = \frac{1}{n}\sum Z_iZ_i'(\hat e_i^{(s)})^2$
\item Re-estimate: $\hat\beta^{(s+1)} = (X'Z\hat\Omega^{(s)-1}Z'X)^{-1}(X'Z\hat\Omega^{(s)-1}Z'Y)$
\item Repeat until convergence
\end{enumerate}

\medskip
In R's \texttt{gmm} package: \texttt{type = "twoStep"} (two-step) or \texttt{type = "iterative"} (iterated).

\medskip
\textbf{Note:} Hansen and Lee (2021) show the iterated GMM estimator is unaffected by whether the weight matrix is computed with or without centering.

\end{frame}

%--- Slide 16: Covariance Matrix Estimation ---
\begin{frame}{Covariance Matrix Estimation}

\textbf{One-step or two-step GMM} (general $W$):
$$\hat V_\beta = (\hat Q'\hat W\hat Q)^{-1}(\hat Q'\hat W\hat\Omega\hat W\hat Q)(\hat Q'\hat W\hat Q)^{-1}$$
where $\hat Q = \frac{1}{n}\sum Z_iX_i'$ and $\hat\Omega$ uses residuals from the final step.

\medskip
\textbf{Efficient iterated GMM} ($\hat W = \hat\Omega^{-1}$):
$$\hat V_\beta = \left(\hat Q'\hat\Omega^{-1}\hat Q\right)^{-1}$$
The sandwich simplifies --- just like GLS vs.\ OLS variance formulas.

\medskip
\begin{tcolorbox}[colback=blue!5, colframe=blue!50]
\textbf{Practical advice:} Always report robust (sandwich) standard errors for one-step GMM. For efficient GMM, the simplified formula is already robust.
\end{tcolorbox}

\end{frame}

%========================================================
% SECTION 4: MISSING DATA APPLICATION
%========================================================
\section{Application: Missing Data}

%--- Slide 17: Missing Data ---
\begin{frame}{Application: Missing Data}

Missing data on regressors is pervasive.  Abrevaya \& Donald (2017) surveyed top journals and found $\sim$50\% of empirical papers have missing data; $\sim$70\% simply drop incomplete observations.

\medskip\pause
\textbf{Standard approaches:}
\begin{enumerate}
\item \textbf{Complete case:} Drop missing obs.---consistent, but loses efficiency
\item \textbf{Dummy variable:} Fill in 0, add indicator---\alert{can be inconsistent}
\item \textbf{Linear imputation:} Predict missing values---requires extra assumptions
\end{enumerate}

\medskip\pause
The GMM approach: write down moment conditions for \emph{both} complete and incomplete observations.  The complete cases contribute structural moments; the incomplete cases contribute reduced-form moments.  GMM combines them optimally.

\end{frame}

%--- Slide 18: The Missing Data Model (Abrevaya 2017) ---
\begin{frame}{The Missing Data Model (Abrevaya \& Donald, 2017)}

\textbf{Structural regression:}
$$y_i = \alpha_0 x_i + z_i'\beta_0 + \varepsilon_i \qquad \text{where } E(x_i\varepsilon_i) = 0,\; E(z_i\varepsilon_i) = 0$$

$x_i$ is a (possibly missing) scalar regressor; $z_i$ is always observed.

\medskip
\textbf{Linear projection of $x_i$ on $z_i$:}
$x_i = z_i'\gamma_0 + \xi_i$ where $E(z_i\xi_i) = 0$.
\textbf{Missingness indicator:} $m_i = 1$ if $x_i$ missing, $m_i = 0$ if observed.

\medskip
\textbf{Key substitution:} For observations with $m_i = 1$:
$$y_i = z_i'(\gamma_0\alpha_0 + \beta_0) + (\varepsilon_i + \xi_i\alpha_0) \equiv z_i'\delta_0 + \eta_i$$

\medskip
\textbf{Assumption 1:} (a) $E(m_iz_i\varepsilon_i) = 0$; (b) $E(m_iz_i\xi_i) = 0$; (c) $E(m_ix_i\varepsilon_i) = 0$.

$\Rightarrow$ Missingness may depend on $z_i$ but not on $\varepsilon_i$ or $\xi_i$.
\end{frame}

%--- Slide 19: Three Blocks of Moment Conditions ---
\begin{frame}{Three Blocks of Moment Conditions}

The parameters are $\theta = (\alpha_0, \beta_0, \gamma_0)'$ with $k = 2K+1$ unknowns.

\medskip
\textbf{Block 1} (observed cases, structural equation):
$(1-m_i)\cdot w_i(y_i - \alpha_0 x_i - z_i'\beta_0) = 0$ \hfill [$K+1$ conditions]

\textbf{Block 2} (observed cases, projection equation):
$(1-m_i)\cdot z_i(x_i - z_i'\gamma_0) = 0$ \hfill [$K$ conditions]

\textbf{Block 3} (missing cases, substituted equation):
$m_i \cdot z_i(y_i - z_i'(\gamma_0\alpha_0 + \beta_0)) = 0$ \hfill [$K$ conditions]

\medskip
\textbf{Total:} $\ell = 3K + 1$ moments for $k = 2K + 1$ parameters $\Rightarrow$ $K$ \alert{overidentifying restrictions} (testable via J-test!)

\smallskip
With $K = 2$ (intercept + IQ): 7 moments, 5 parameters, \textbf{2 overidentifying restrictions}

\end{frame}

%--- Slide 20: Why GMM Beats the Alternatives ---
\begin{frame}{Why GMM Beats the Alternatives}

\vspace{-4pt}
{\footnotesize
\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Consistent?} & \textbf{Efficient?} & \textbf{Testable?} \\
\midrule
Complete case & Yes (drops $m_i\!=\!1$) & No (fewer obs) & No \\
Dummy variable & \alert{Sometimes no} & No & No \\
Linear imputation & Yes (under Assn 1) & Somewhat & No \\
\textbf{GMM} & \textbf{Yes} (under Assn 1) & \textbf{Yes} (optimal $W$) & \textbf{Yes} (J-test) \\
\bottomrule
\end{tabular}
\end{table}
}

\smallskip
\textbf{Abrevaya \& Donald Monte Carlo} ($n=400$, 1000 replications):

{\footnotesize
\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Bias ($\alpha_0$)} & \textbf{$n \times$Var ($\alpha_0$)} & \textbf{MSE ($\alpha_0$)} \\
\midrule
Complete case & 0.011 & 13.93 & 0.035 \\
Dummy variable & $-0.194$ & 13.94 & 0.073 \\
FGLS & 0.011 & 13.93 & 0.035 \\
\textbf{GMM} & \textbf{0.006} & \textbf{12.53} & \textbf{0.031} \\
\bottomrule
\end{tabular}
\end{table}
}

\end{frame}

%--- Slide 21: WLS Data (R code) ---
\begin{frame}[fragile]{WLS Data: Education and BMI}

\begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
library(gmm); library(haven); library(estimatr)

# Wisconsin Longitudinal Study data
data1 <- read_dta("wls-data.dta")
data_men <- subset(data1, male == 1)

# How much data is missing?
table(data_men$bmimissing)
#  0    1
# 4231 1095   (~21% missing)

# Complete-case regression
data_complete <- subset(data_men, bmimissing == 0)
cc_reg <- lm(educ ~ bmirating + iq, data = data_complete)
\end{lstlisting}
Education ($y$) regressed on BMI rating ($x$, sometimes missing) and IQ ($z$, always observed).

\end{frame}

%--- Slide 22: Defining GMM Moment Functions in R ---
\begin{frame}[fragile]{Defining GMM Moment Functions in R}

\begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
g_men <- function(theta, x) {
  beta_0 <- theta[1]; alpha <- theta[2]; beta_iq <- theta[3]
  gamma_0 <- theta[4]; gamma_iq <- theta[5]
  educ <- x[,1]; bmirating <- x[,2]; iq <- x[,3]; bmimissing <- x[,4]

  # Block 1: structural eq (observed cases)
  r1 <- (1 - bmimissing) * (educ - beta_0 - alpha*bmirating - beta_iq*iq)
  # Block 2: projection eq (observed cases)
  r2 <- (1 - bmimissing) * (bmirating - gamma_0 - gamma_iq*iq)
  # Block 3: substituted eq (missing cases)
  r3 <- bmimissing * (educ - (gamma_0*alpha+beta_0) - (gamma_iq*alpha+beta_iq)*iq)

  cbind(r1, r1*bmirating, r1*iq,  # 3 moments
        r2, r2*iq,                  # 2 moments
        r3, r3*iq)                  # 2 moments = 7 total, 5 params
}
\end{lstlisting}

\end{frame}

%--- Slide 23: Running GMM in R ---
\begin{frame}[fragile]{Running GMM in R}

\begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
# Starting values from complete-case regressions
start_men <- c(beta_0 = coef(cc_reg)[1],
               alpha  = coef(cc_reg)[2],
               beta_iq = coef(cc_reg)[3],
               gamma_0 = coef(proj_reg)[1],
               gamma_iq = coef(proj_reg)[2])

# Two-step GMM with HAC variance
gmm_men <- gmm(g_men, x = x_men, t0 = start_men,
                type = "twoStep",
                wmatrix = "ident", vcov = "HAC")
summary(gmm_men)
\end{lstlisting}
\texttt{type = "twoStep"}: initial $W = I$, then update with $\hat\Omega^{-1}$.  \texttt{vcov = "HAC"}: HAC variance estimator.

\end{frame}

%--- Slide 24: Comparing Methods: Results ---
\begin{frame}[fragile]{Comparing Methods: Results}

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
& \textbf{Complete Case} & \textbf{Dummy Variable} & \textbf{GMM} \\
\midrule
$\hat\alpha$ (BMI) & $-0.0343$ & $-0.0343$ & $-0.0340$ \\
SE & $(0.0077)$ & $(0.0073)$ & $(0.0072)$ \\[4pt]
$\hat\beta_{IQ}$ & $0.0563$ & $0.0567$ & $0.0564$ \\
SE & $(0.0017)$ & $(0.0015)$ & $(0.0015)$ \\[4pt]
$n$ used & 4231 & 5326 & 5326 \\
\bottomrule
\end{tabular}
\end{table}

\medskip
\textbf{J-test for overidentifying restrictions:}
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
specTest(gmm_men)  # J-test = ?, df = 2, p-value = ?
\end{lstlisting}

\begin{itemize}
\item Large p-value: fail to reject $\Rightarrow$ Assumption 1 is plausible
\item Small p-value: evidence against the linear projection restriction
\end{itemize}

\end{frame}

%========================================================
% SECTION 5: NONLINEAR GMM AND SUMMARY
%========================================================
\section{Nonlinear GMM and Summary}

%--- Slide 25: Nonlinear GMM ---
\begin{frame}{Nonlinear GMM}

So far: $g_i(\beta)$ is linear in $\beta$. In general, $g_i(\beta)$ can be \textbf{nonlinear}.

\medskip
The GMM estimator still minimizes $J(\beta) = n\,\bar{g}_n(\beta)'\hat W\,\bar{g}_n(\beta)$ but now requires numerical optimization (no closed form).

\medskip
\begin{block}{Proposition 13.1: Distribution of Nonlinear GMM}
$\sqrt{n}(\hat\beta_{\text{gmm}} - \beta) \xrightarrow{d} N(0, V_\beta)$
where $V_\beta = (Q'WQ)^{-1}(Q'W\Omega WQ)(Q'WQ)^{-1}$ with the \textbf{Jacobian}
$Q = \mathbb{E}\!\left[\frac{\partial}{\partial \beta'} g_i(\beta)\right]$
replacing the linear $\mathbb{E}[ZX']$.
\end{block}

\smallskip
\textbf{Example:} The Abrevaya \& Donald missing data model is nonlinear --- Block 3 contains $\gamma_0\alpha_0$ (product of parameters).

\end{frame}

%--- Slide 26: The MTE Connection ---
\begin{frame}{The MTE Connection}

Recall from Lecture 14: the \textbf{Marginal Treatment Effect} is
$$\Delta^{MTE}(x, u_D) = \mathbb{E}[Y_1 - Y_0 \mid X = x, U_D = u_D]$$

\medskip
\textbf{MTE estimation is nonlinear GMM:}
\begin{itemize}\itemsep=1pt
\item Moment conditions from the relationship between $E[Y \mid X, Z]$ and propensity score $P(Z)$
\item MTE function modeled as a polynomial in $u_D$; parameters enter nonlinearly
\item Overidentification arises from multiple instruments $\Rightarrow$ $J$-test applies
\end{itemize}

\medskip
The \texttt{ivmte} package (Shea \& Torgovitsky) implements this: specifying moment conditions, choosing weight matrices, and testing overidentifying restrictions.

\end{frame}

%--- Slide 27: Summary of GMM Estimation (Flowchart) ---
\begin{frame}{Summary of GMM Estimation}

\begin{center}
\begin{tikzpicture}[
  node distance=1.0cm and 0.5cm,
  box/.style={rectangle, draw=blue!50, fill=blue!5, text width=3.2cm, minimum height=0.9cm, align=center, font=\small},
  arrow/.style={-{Stealth[length=3mm]}, thick, blue!70}
]

\node[box] (specify) {1. Specify moment conditions\\$\mathbb{E}[g_i(\beta)] = 0$};
\node[box, right=of specify] (choose) {2. Choose initial $W$\\(e.g., $I_\ell$ or $(Z'Z)^{-1}$)};
\node[box, right=of choose] (minimize) {3. Minimize\\$J(\beta) = n\bar{g}_n'W\bar{g}_n$};
\node[box, below=of minimize] (residuals) {4. Compute residuals\\$\hat e_i = Y_i - X_i'\hat\beta$};
\node[box, left=of residuals] (update) {5. Update weight\\$\hat W = \hat\Omega^{-1}$};
\node[box, left=of update] (converge) {6. Converge?\\Report $\hat\beta$, $\hat V_\beta$, $J$};

\draw[arrow] (specify) -- (choose);
\draw[arrow] (choose) -- (minimize);
\draw[arrow] (minimize) -- (residuals);
\draw[arrow] (residuals) -- (update);
\draw[arrow] (update) -- (converge);
\draw[arrow, dashed, darkpurple] (update.north) -- ++(0,0.6) -| (minimize.south) node[pos=0.25, above, font=\footnotesize\color{darkpurple}] {iterate};

\end{tikzpicture}
\end{center}

\smallskip
\begin{itemize}
\item \textbf{One-step:} Stop after step 3 (e.g., 2SLS with $W=(Z'Z)^{-1}$)
\item \textbf{Two-step:} One pass through steps 3--5
\item \textbf{Iterated:} Repeat steps 3--5 until convergence
\end{itemize}

\end{frame}

%--- Slide 28: The OLS → GLS Parallel ---
\begin{frame}{The OLS $\to$ GLS Parallel}

The progression from 2SLS to efficient GMM mirrors Lectures 5--6:

\smallskip
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{lll}
\toprule
& \textbf{Linear model} & \textbf{IV model} \\
\midrule
Baseline & OLS & 2SLS \\
Efficient under homoskedasticity? & Yes & Yes \\
Efficient under heteroskedasticity? & No $\Rightarrow$ GLS & No $\Rightarrow$ efficient GMM \\
Variance formula & $(X'\Omega^{-1}X)^{-1}$ & $(Q'\Omega^{-1}Q)^{-1}$ \\
Estimated how? & FGLS (two-step) & Two-step GMM \\
\bottomrule
\end{tabular}
\renewcommand{\arraystretch}{1.0}

\medskip\pause
GMM is to IV what GLS is to OLS: use the covariance structure of the moments to improve efficiency.

\end{frame}

%--- Slide 29: Where We Stand ---
\begin{frame}{Where We Stand}

\begin{tcolorbox}[colback=blue!5, colframe=blue!50, title=What GMM Buys You]
\begin{enumerate}\itemsep=1pt
\item \textbf{No distributional assumptions.}  You specify moment conditions, not a likelihood---only orthogonality conditions.
\item \textbf{Optimal weighting.}  $W = \Omega^{-1}$ gives the smallest asymptotic variance, paralleling GLS.
\item \textbf{Testable restrictions.}  Extra moment conditions generate overidentifying restrictions you can test with the $J$-statistic (next lecture).
\end{enumerate}
\end{tcolorbox}

Every estimator in this course---OLS, GLS, IV, 2SLS---is a special case.  GMM makes explicit what each one assumes and what it trades off.

\end{frame}

%--- Slide 30: Preview of Next Lecture ---
\begin{frame}{Preview of Next Lecture}

\textbf{Lecture 16: GMM Inference and Going Beyond ATE}

\medskip
\begin{itemize}
\item \textbf{Testing:} The J-test for overidentification (Thm 13.14)
\item \textbf{Hypothesis tests:} Wald test and Distance test in the GMM framework
\item \textbf{Subset tests:} Testing specific instruments, endogeneity tests
\item \textbf{Application: Propaganda and Foreign Media}
\begin{itemize}
\item Kern \& Hainmueller: West German TV in East Germany
\item IV gives LATE $\approx -0.12$ (media reduces support for regime)
\item MTE reveals \alert{treatment effect heterogeneity}: always-takers vs.\ never-takers have \textit{opposite signs}
\end{itemize}
\item \textbf{The course in one slide:} OLS $\to$ GLS $\to$ IV $\to$ 2SLS $\to$ GMM
\end{itemize}

\end{frame}

\end{document}
